Factors affecting reliablity of state-space age-structured assessment models

### 2025-02-21 Project_0/manuscript/manuscript.tex
saving work on project_0 manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Brandon
Muffley\textsuperscript{5} \and Brian C.
Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{12 February, 2025}

@@

Much is known about the reliability of state-space models that are
linear or Gaussian (Aeberhard et al. 2018), but applications in
fisheries management are nonlinear and typically include multiple types
of observations with varying distributional assumptions. Furthermore,
there is a wide range of potential random effects structures and model
parameters that can be treated as random effects in assessment models.
We know relatively little about the factors affecting statistical
reliability of such models or the ability of information criteria to
distinguish among such alternative structures.

Li et al. (2024) investigated some aspects of inferences for operating
models with multiple sources of process error, but there are differences
for this paper.

Here we conduct a simulation study with operating models (OMs) varying
by degree of observation error uncertainty, source and variability of
process error, and fishing history. The simulations from these OMs are
fitted with estimation models (EMs) that make alternative assumptions
for sources of process error, whether a stock-recruit model was
estimated, and whether M is estimated. We evaluate whether AIC can
correctly determine the correct source of process error and stock
effects on recruitment. We also evaluate the degree of bias in the
outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure operating
and EMs in our simulation study (Miller and Stock 2020; Stock and Miller
2021). WHAM is an R package freely available as a github repository. For
this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This package has also been used to
configure operating and EMs for closed loop simulations evaluating
index-based assessment methods (Legault et al. 2023) and is used for
management of haddock, butterfish, American plaice, bluefish, Atlantic
cod, black sea bass, and yellowtail flounder in the Northeast US.

@@

For R and R+S OMs, convergence generally declines for most EMs when the
median natural mortality rate is estimated and/or the Beverton-Holt
stock recruit relationship is estimated even when the process error
assumptions of the EMs and OMs match (Figure
\ref{hessian_SE_convergence}, A). When there was high observation error
and constant fishing pressure (change from \(2.5F_{\text{MSY}}\) to
\(F_{\text{MSY}}\)), convergence of all EM process error configurations
other than R EMs was poor for R OMs (\(\sigma_{2+} = 0\)) regardless of
whether median natural mortality and stock-recruit relationships were
estimated. Convergence of R EMs was high for all R and R+S OMs except
when there was high observation error and constant fishing pressure and
median natural mortality and stock-recruit relationships were estimated.
R+S EMs fit to R OMs exhibited poor convergence regardless of whether
natural mortality or a stock-recruit relationship was estimated. R+S EMs
fit to R+S OMs had highest convergence rates when there was contrast in
fishing pressure and low observation error. Convergence rates were high
for all EMs when fit to data from R+S OMs with lower observation
uncertainty except those where median natural mortality and/or
stock-recruit relationships were estimated.

@@

With R and R+S process error assumptions, probability of AIC accuracy
for the B-H stock-recruit relationship as a function of SSB variability
were greatest for OMs where there was greater process variability in
survival, lower process variability in recruitment and contrast in
fishing pressure (Figure \ref{sr_aic}, A). However, the largest
variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs
with larger recruitment variability (\(\sigma_R = 1.5\); Figure
\ref{sr_aic}, A, right column group). The weakest overall accuracy of
AIC for the Beverton-Holt stock-recruit function was when there was
constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth
rows). The largest effect of simultaneously estimating median natural
mortality was when there was OMs had high observation uncertainty and
constant fishing pressure (Figure \ref{sr_aic}, A, second row).

For R+M OMs, probability of AIC accuracy for the Beverton-Holt
stock-recruit relationship was greatest for OMs where there was contrast
in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in
SSB occurred in OMs with larger natural mortality variability and
temporal AR1 correlation of the annual random effects
(\(\sigma_M = 1.5\), \(\rho_M = 0.9\); Figure \ref{sr_aic}, B,
right-most column) where AIC accuracy was also high even when there was
constant fishing pressure. Interestingly, there seemed to be little
difference in AIC accuracy whether there was lower or higher observation
uncertainty or whether the EM simultaneously estimated median natural
mortality.

@@

For R OMs (\(\sigma_{2+} = 0\)), there was no indication of bias (95\%
confidence intervals included 0) in terminal year SSB for any of the
estimating models regardless of process error assumptions, except when
no SR assumption was made, recruitment variability was low, and there
was contrast in fishing mortality and high observation uncertainty
(Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB
estimates were highly variable when median natural mortality was
estimated and there was constant fishing pressure and high observation
uncertainty (Figure \ref{SSB_rel_error}, A, second row).

For R+S OMs, the EMs with matching process error assumption generally
produced unbaised estimation of terminal SSB except when median natural
mortality was estimated and there was high observation uncertainty. In
R+S OMs with low observation uncertainty, EMs with incorrect process
error assumptions typically provided biased estimation of terminal year
SSB. Estimating the Beverton-Holt stock-recruit relationship had little
discernible effect on bias of terminal year SSB estimation whereas
estimating median M tended to produce more variability in errors in
terminal SSB estimation similar to R OMs.

For R+ M OMs with low variability in natural mortality process errors,
low observation uncertainty and contrast in fishing motality over time
all EMs produced low variability in SSB estimation error that indicated
unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger
variability in natural mortality process errors increased bias of EMs
without the correct process error type. Estimating median natural
mortality increased variability of SSB estimation error particularly for
OMs with high observation uncertainty and constant fishing pressure over
time and increased bias in SSB estimation for many R+M OMs. Like R and
R+S OMs, estimating a stock-recruit relationship had little discernible
effect on SSB bias.

For R+Sel OMs, there was no evidence of bias for any EMs when
variability in selectivity process errors was low, observation
uncertainty low and contrast in fishing mortality (Figure
\ref{SSB_rel_error}, C). The largest bias for all EMs that estimated
median natural mortality when the OMs had high observation uncertainty,
constant fishing pressure and greater variability in selectivity process
errors (\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
(\(\sigma_{\text{Sel}} = 0.1\)) and low observation uncertainty.
However, there was no evidence of bias of matching R+Sel EMs when
observation uncertainty was low and variation in selectivity process
errors was larger, whether median natural mortality was estimated or not
(Figure \ref{SSB_rel_error}, C, third row). We only observed an effect
of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs
that had high observation uncertainty and contrast in fishing pressure
where estimating the relationship produced less biased SSB estimation
for many EMs (Figure \ref{SSB_rel_error}, C, top row).

All EMs fit to data from R+q OMs with low observation uncertainty and
contrast in fishing pressure exhibited little evidence of bias in
terminal SSB estimation except for R+M EMs when there was no AR1
correlation in catchability process errors (Figure \ref{SSB_rel_error},
D). Many EMs also performed well in R+q OMs with low observation
uncertainty, but no contrast in fishing pressure. For R+q OMs with high
observation uncertainty and contrast in fishing pressure, EMs that
estimated the Beverton-Holt stock recruit function exhibited less SSB
bias than those that did not. Estimating median natural mortality in the
EMs only resulted in much more variable SSB estimation errors when there
was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D,
first and third rows).

@@

Median natural mortality rate is estimated accurately by all EM process
error types for all R OMs except those where observation uncertainty is
high and fishing pressure is constant where variability in relative
errors is high (Figure \ref{M_rel_error}, A, \(\sigma_{2+} = 0\)). For
R+S OMs estimation of median natural mortality rate is most accurate
when observation uncertainty is low and there is contrast in fishing
pressure and the EM process error type is correct. There is little
effect of estimating stock-recruit relationships on the bias for median
natural mortality.

For R+M OMs, median natural mortality was estimated most accurately,
regardless of EM process error type, when OMs had a change in fishing
pressure and low observation uncertainty (Figure \ref{M_rel_error}, B).
However, those R+M OMs that also had greatest variability in AR1
correlated natural mortality process errors only had unbiased estimation
when the EM process error type was correct. Like R, and R+S OMs, there
was little effect of estimating stock-recruit relationships on the bias
for median natural mortality.

All EM process error types produced accurate estimation of median
natural mortality rate for R+Sel OMs that had contrast in fishing
pressure, low observation uncertainty, and low selectivity process error
variability (Figure \ref{M_rel_error}, C). When selectivity process
error variability increased the incorrect EM process errors produce more
biased estimation of median natural mortality rate. The least accurate
estimation occurred for all EM process error types when observation
uncertainty was high and there was constant fishing pressure. Again,
like other OM process error types, there was little effect of estimating
stock-recruit relationships on the bias for median natural mortality.

Like R+Sel OMs, all EM process error types produced accurate estimation
of median natural mortality rate when fit to R+q OMs with contrast in
fishing pressure, low observation uncertainty and low catchability
process error variability (Figure \ref{M_rel_error}, D). Most EM process
error types produced biased estimation of median natural mortality when
R+q OMs had high observaiton uncertainty and constant fishing pressure.
Again, there was no discernible effect of estimating stock-recruit
relationships on bias.

@@

Mohn's \(\rho\) for SSB was small in absolute value for all R and R+S
OMs, regardless of EM process error types, and whether median natural
mortality rate or stock-recruit relationships were estimated (Figure
\ref{mohns_rho_ssb}, A). The strongest retrospective patterns occurred
in OMs with the largest survival process error variability, high
observation uncertainty, and contrast in fishing pressure, but only for
EMs without the correct process error type and where median natural
mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's
\(\rho\) was also small in absolute value, but median values were all
closer to 0 than the largest values in the R and R+S OMs (Figure
\ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect
of estimation of median natural mortality rate or stock recruit
relationship on Mohn's \(\rho\) for any EM process error types.

Mohn's \(\rho\) for recruitment was small in absolute value for all R
OMs with low variability in recruitment process errors, regardless of EM
process error type, and whether median natural mortality rate or
stock-recruit relationships were estimated (Figure \ref{mohns_rho_R},
A). However, R and R+S OMs with greater recruitment process variability
and higher observation uncertainty had median Mohn's \(\rho\) for
recruitment greater than zero for most EMs even when the the EM process
error type was correct. In R+S OMs with lower observation uncertainty,
EMs with the correct process error type exhibited better median Mohn's
\(\rho\) close to 0 than EMs with the incorrect process error type. For
R+M, R+Sel, and R+q OMs, results for Mohn's \(\rho\) for recruitment are
similar to those for SSB, but the range in median values and variation
in Mohn's \(\rho\) values for a given OM are generally larger for
recruitment (Figure \ref{mohns_rho_R}, B-D).

@@

Convergence results can be useful for understanding how lack of
convergence in applications to real data might direct the practitioner
to which alternative random effects configurations is more appropriate.
Therefore, the type of convergence that we might use as a diagnostic is
important. It is common during assessment model fitting to check that
the maximum absolute gradient component is less than some threshold, but
there is no standard. However, checking the Hessian for positive
definiteness is sometimes done conditional on the convergence criterion.
As such, some models may be excluded that in fact have an invertible
hessian. We found the hessian at the optimized log-likelihood can often
be invertible when the gradient approach with a sensible criterion
(e.g., \(10^{-3}\)) would indicate lack of convergence.

Another factor affecting the convercence criteria is numerical accuracy.
With the process errors modeled as random effects we are optimizing an
(Laplace) approximation of the marginal log-likelihood and therefore
hessians and gradients are also with regard to the approximation.
Furthermore, numerical methods are used to calculate the hessian
(Optimhess?), and also to invert it for variance estimation. All of
these approximations are affecting the gradients and invertibility of
the hessian (although likely worse for some models than others), which,
along with our results, suggests at least investigating whether the
hessian is positive definite when the calculated gradients are not
terribly large.

Often high gradients and/or lack of positive definite hessian result
when selectivity parameter estimates are at their bounds (0,1) or random
effects variance parameter estimates are essentially 0. Changing the
parameter configuration to avoid the bounds or removing process errors
with 0 variance will often lead to satisfactory convergence.

Perhaps a greater gradient threshold would be more consistent with
hessian-based convergence, but the gradient and hessian-based standard
errors should be inspected.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{13 February, 2025}

@@

Although the statistical aspects of state-space models and their
application have been studied extensively, the work is primarily on
Gaussian state-space models. State-space stock assessment models, where
there are nonlinear functions of latent parameters and multiple types of
observations with varying distributional assumptions, is one of the most
complex examples of this analytical approach.

Much is known about the reliability of state-space models that are
linear or Gaussian (Aeberhard et al. 2018), but not for the more complex
application in stock assessment modeling.

(Li et al. 2024 investigated some aspects of inferences for operating
models with multiple sources of process error, but there are differences
for this paper.)

There is a wide range of model parameters that can be treated as random
effects in assessment models and a variety of potential distributions
for the random effects. We know relatively little about the factors
affecting statistical reliability of such models or the ability to
distinguish among such alternative structures.

Here we conduct a simulation study with operating models (OMs) varying
by degree of observation error error, source and variability of process
error, and fishing history. The simulations from these OMs are fitted
with estimation models (EMs) that make alternative assumptions for
sources of process error, whether a stock-recruit model was estimated,
and whether a constant, or, in some EMs, median, natural mortality is
estimated. We evaluate whether AIC can correctly determine the correct
source of process error and a stock-recruit relationship. We also
evaluate when retrospective patterns occur and the degree of bias in the
outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
WHAM is an R package freely available as a github repository. For this
study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This package has also been used to
configure OMs and EMs for closed loop simulations evaluating index-based
assessment methods (Legault et al. 2023) and is used for management of
haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea
bass, and yellowtail flounder in the Northeast US.

@@

For R and R+S OMs, convergence generally declines for most EMs when the
median natural mortality rate is estimated and/or the Beverton-Holt
stock recruit relationship is estimated even when the process error
assumptions of the EMs and OMs match (Figure
\ref{hessian_SE_convergence}, A). When there was high observation error
and constant fishing pressure (change from \(2.5F_{\text{MSY}}\) to
\(F_{\text{MSY}}\)), convergence of all EM process error configurations
other than R EMs was poor for R OMs (\(\sigma_{2+} = 0\)) regardless of
whether median natural mortality and stock-recruit relationships were
estimated. Convergence of R EMs was high for all R and R+S OMs except
when there was high observation error and constant fishing pressure and
median natural mortality and stock-recruit relationships were estimated.
R+S EMs fit to R OMs exhibited poor convergence regardless of whether
natural mortality or a stock-recruit relationship was estimated. R+S EMs
fit to R+S OMs had highest convergence rates when there was contrast in
fishing pressure and low observation error. Convergence rates were high
for all EMs when fit to data from R+S OMs with lower observation error
except those where median natural mortality and/or stock-recruit
relationships were estimated.

@@

With R and R+S process error assumptions, probability of AIC accuracy
for the B-H stock-recruit relationship as a function of SSB variability
were greatest for OMs where there was greater process variability in
survival, lower process variability in recruitment and contrast in
fishing pressure (Figure \ref{sr_aic}, A). However, the largest
variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs
with larger recruitment variability (\(\sigma_R = 1.5\); Figure
\ref{sr_aic}, A, right column group). The weakest overall accuracy of
AIC for the Beverton-Holt stock-recruit function was when there was
constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth
rows). The largest effect of simultaneously estimating median natural
mortality was when there was OMs had high observation error and constant
fishing pressure (Figure \ref{sr_aic}, A, second row).

For R+M OMs, probability of AIC accuracy for the Beverton-Holt
stock-recruit relationship was greatest for OMs where there was contrast
in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in
SSB occurred in OMs with larger natural mortality variability and
temporal AR1 correlation of the annual random effects
(\(\sigma_M = 1.5\), \(\rho_M = 0.9\); Figure \ref{sr_aic}, B,
right-most column) where AIC accuracy was also high even when there was
constant fishing pressure. Interestingly, there seemed to be little
difference in AIC accuracy whether there was lower or higher observation
error or whether the EM simultaneously estimated median natural
mortality.

@@

For R OMs (\(\sigma_{2+} = 0\)), there was no indication of bias (95\%
confidence intervals included 0) in terminal year SSB for any of the
estimating models regardless of process error assumptions, except when
no SR assumption was made, recruitment variability was low, and there
was contrast in fishing mortality and high observation error (Figure
\ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were
highly variable when median natural mortality was estimated and there
was constant fishing pressure and high observation error (Figure
\ref{SSB_rel_error}, A, second row).

For R+S OMs, the EMs with matching process error assumption generally
produced unbaised estimation of terminal SSB except when median natural
mortality was estimated and there was high observation error. In R+S OMs
with low observation error, EMs with incorrect process error assumptions
typically provided biased estimation of terminal year SSB. Estimating
the Beverton-Holt stock-recruit relationship had little discernible
effect on bias of terminal year SSB estimation whereas estimating median
M tended to produce more variability in errors in terminal SSB
estimation similar to R OMs.

For R+ M OMs with low variability in natural mortality process errors,
low observation error and contrast in fishing motality over time all EMs
produced low variability in SSB estimation error that indicated
unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger
variability in natural mortality process errors increased bias of EMs
without the correct process error type. Estimating median natural
mortality increased variability of SSB estimation error particularly for
OMs with high observation error and constant fishing pressure over time
and increased bias in SSB estimation for many R+M OMs. Like R and R+S
OMs, estimating a stock-recruit relationship had little discernible
effect on SSB bias.

For R+Sel OMs, there was no evidence of bias for any EMs when
variability in selectivity process error and observation error was low,
and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C).
The largest bias occurred for any EMs that estimated median natural
mortality when the OMs had high observation error, constant fishing
pressure, and greater variability in selectivity process errors
(\(\sigma_{\text{Sel}} = 0.5\)) or low selectivity process errors
(\(\sigma_{\text{Sel}} = 0.1\)) and low observation error. However,
there was no evidence of bias of matching R+Sel EMs when observation
error was low and variation in selectivity process errors was larger,
whether median natural mortality was estimated or not (Figure
\ref{SSB_rel_error}, C, third row). We only observed an effect of
estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs
that had high observation error and contrast in fishing pressure where
estimating the relationship produced less biased SSB estimation for many
EMs (Figure \ref{SSB_rel_error}, C, top row).

All EMs fit to data from R+q OMs with low observation error and contrast
in fishing pressure exhibited little evidence of bias in terminal SSB
estimation except for R+M EMs when there was no AR1 correlation in
catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs
also performed well in R+q OMs with low observation error, but no
contrast in fishing pressure. For R+q OMs with high observation error
and contrast in fishing pressure, EMs that estimated the Beverton-Holt
stock recruit function exhibited less SSB bias than those that did not.
Estimating median natural mortality in the EMs only resulted in much
more variable SSB estimation errors when there was no contrast in
fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).

@@

Median natural mortality rate is estimated accurately by all EM process
error types for all R OMs except those where observation error is high
and fishing pressure is constant where variability in relative errors is
high (Figure \ref{M_rel_error}, A, \(\sigma_{2+} = 0\)). For R+S OMs
estimation of median natural mortality rate is most accurate when
observation error is low and there is contrast in fishing pressure and
the EM process error type is correct. There is little effect of
estimating stock-recruit relationships on the bias for median natural
mortality.

For R+M OMs, median natural mortality was estimated most accurately,
regardless of EM process error type, when OMs had a change in fishing
pressure and low observation error (Figure \ref{M_rel_error}, B).
However, those R+M OMs that also had greatest variability in AR1
correlated natural mortality process errors only had unbiased estimation
when the EM process error type was correct. Like R, and R+S OMs, there
was little effect of estimating stock-recruit relationships on the bias
for median natural mortality.

All EM process error types produced accurate estimation of median
natural mortality rate for R+Sel OMs that had contrast in fishing
pressure, low observation error, and low selectivity process error
variability (Figure \ref{M_rel_error}, C). When selectivity process
error variability increased the incorrect EM process errors produce more
biased estimation of median natural mortality rate. The least accurate
estimation occurred for all EM process error types when observation
error was high and there was constant fishing pressure. Again, like
other OM process error types, there was little effect of estimating
stock-recruit relationships on the bias for median natural mortality.

Like R+Sel OMs, all EM process error types produced accurate estimation
of median natural mortality rate when fit to R+q OMs with contrast in
fishing pressure, low observation error and low catchability process
error variability (Figure \ref{M_rel_error}, D). Most EM process error
types produced biased estimation of median natural mortality when R+q
OMs had high observaiton error and constant fishing pressure. Again,
there was no discernible effect of estimating stock-recruit
relationships on bias.

@@

Mohn's \(\rho\) for SSB was small in absolute value for all R and R+S
OMs, regardless of EM process error types, and whether median natural
mortality rate or stock-recruit relationships were estimated (Figure
\ref{mohns_rho_ssb}, A). The strongest retrospective patterns occurred
in OMs with the largest survival process error variability, high
observation error, and contrast in fishing pressure, but only for EMs
without the correct process error type and where median natural
mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's
\(\rho\) was also small in absolute value, but median values were all
closer to 0 than the largest values in the R and R+S OMs (Figure
\ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect
of estimation of median natural mortality rate or stock recruit
relationship on Mohn's \(\rho\) for any EM process error types.

Mohn's \(\rho\) for recruitment was small in absolute value for all R
OMs with low variability in recruitment process errors, regardless of EM
process error type, and whether median natural mortality rate or
stock-recruit relationships were estimated (Figure \ref{mohns_rho_R},
A). However, R and R+S OMs with greater recruitment process variability
and higher observation error had median Mohn's \(\rho\) for recruitment
greater than zero for most EMs even when the the EM process error type
was correct. In R+S OMs with lower observation error, EMs with the
correct process error type exhibited better median Mohn's \(\rho\) close
to 0 than EMs with the incorrect process error type. For R+M, R+Sel, and
R+q OMs, results for Mohn's \(\rho\) for recruitment are similar to
those for SSB, but the range in median values and variation in Mohn's
\(\rho\) values for a given OM are generally larger for recruitment
(Figure \ref{mohns_rho_R}, B-D).

@@

Convergence results can be useful for understanding how lack of
convergence in applications to real data might direct the practitioner
to which alternative random effects configurations is more appropriate.
Therefore, the type of convergence that we might use as a diagnostic is
important. It is common during assessment model fitting to check that
the maximum absolute gradient component is less than some threshold, but
there is no standard. However, if the Hessian were checked for positive
definiteness conditional on a satisfactory gradient, some models may be
excluded that in fact have an invertible hessian. We found the hessian
at the optimized log-likelihood can often be invertible when the
gradient approach with a sensible criterion (e.g., \(10^{-3}\)) would
indicate lack of convergence.

Another factor affecting the convergence criteria is numerical accuracy.
With the process errors modeled as random effects we are optimizing an
(Laplace) approximation of the marginal log-likelihood and therefore
hessians and gradients are also with regard to the approximation.
Furthermore, numerical methods are used to calculate the hessian
(Optimhess?), and also to invert it for variance estimation. All of
these approximations are affecting the gradients and invertibility of
the hessian (although likely worse for some models than others), which,
along with our results, suggests at least investigating whether the
hessian is positive definite when the calculated gradients are not
terribly large. Often high gradients and/or lack of positive definite
hessian result when parameter estimates are at their bounds (Carvalho et
al. 2021), and this also applies to variance or autocorrelation
parameters for random effects with state-space assessment models.
